---
title: "Direct, total and separable effects, using G-formula"
output: 
  rmdformats::readthedown:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Hmisc)
library(splines)
library(here)
source(here::here("R", "utility_functions.R"))

prostate <- read.csv(here::here("R", "prostate.csv"), sep = ';')
prostate$allCause <- prostate$status != "alive"
prostate$eventType <- as.factor(prostate$status)

levels(prostate$eventType) <-
  list(alive = "alive",
       pdeath = "dead - prostatic ca",
       odeath = c(levels(prostate$eventType)[c(2:5, 7:10)]))

prostate$rx <- as.factor(prostate$rx)
prostRed <- prostate[prostate$rx %in% levels(prostate$rx)[3:4], ]
prostRed$temprx = as.integer(prostRed$rx) - 3
prostRed$rx = abs(1 - prostRed$temprx) # DES is A=1 and placebo A=0
prostRed$eventType = as.integer(prostRed$eventType) - 1 #0 is censoring, 1 is pdeath, 2 is odeath
prostRed$hgBinary <- prostRed$hg < 12
prostRed$ageCat <- cut2(prostRed$age, c(0, 60, 70, 80, 100))
prostRed$normalAct <- prostRed$pf == "normal activity"
prostRed$eventCens <- prostRed$eventType == 0
prostRed$eventProst <- prostRed$eventType == 1
prostRed$Tstart = -0.01 #Needed for the long format
cutTimes <- c(0:59)

longProstRed <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "allCause")

longProstRedCens <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "eventCens")

longProstRed$eventCens <- longProstRedCens$eventCens

longProstRed$prostateDeath <-
  longProstRed$allCause == 1 &  longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2

longProstRed$prostateDeath[longProstRed$eventCens == 1] <- NA

longProstRed$otherDeath[longProstRed$eventCens == 1] <- NA

longProstRed$prostateDeath[longProstRed$otherDeath == 1] <- NA

longProstRed <- longProstRed[longProstRed$dtime < length(cutTimes), ]

baseline <- longProstRed[longProstRed$dtime == 0, ]

n <- length(unique(longProstRed$patno))

```


## Create expanded datasets

Create expanded data so that hazards given A=1 can be computed for all subjects (even those with A=0). In an RCT this is not necessary, could stratify models above and this step by treatment arm.

Expand baseline so it contains a visit at each time point for every individual where the baseline information has been carried forward at each time.


```{r}
### Pooled logistic regression for prostate cancer death

plrFitP <-
  glm(
    prostateDeath ~ rx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary ,
    data = longProstRed,
    family = binomial())

### Pooled logistic regression for death due to other causes

plrFitO <-
  glm(
    otherDeath ~ dtime + I(dtime ^ 2) + normalAct + ageCat + hx + hgBinary + rx ,
    data = longProstRed,
    family = binomial())

```

## Direct effect

Dataset construction where A = 1
```{r}
treated <- baseline[rep(1:n,each=length(cutTimes)),] #One row for each interval k+1 for k=0,...,K
treated$dtime <- rep(cutTimes,n)
treated$rx <-1 
```

Dataset construction where A = 0
```{r}
# Make analogous dataset for placebo
placebo <- baseline[rep(1:n,each=length(cutTimes)),] #One row for each time
placebo$dtime <- rep(cutTimes,n)
placebo$rx <- 0 
```

Estimate conditional discrete hazards (cause specific for each cause of death/competing event conditioned for event and hazard of competing event) for each subject in each time interval in both cloned datasets

```{r}
treated$hazardP <- predict(plrFitP, newdata = treated, type = 'response') 
treated$hazardO <- 0
treated$s <- (1-treated$hazardP) * (1-treated$hazardO)

placebo$hazardP <- predict(plrFitP, newdata = placebo, type = 'response') 
placebo$hazardO <- 0
placebo$s <- (1-placebo$hazardP) * (1-placebo$hazardO)
```

Parametric g-formula estimate of risk under elimination of competing events

```{r}
cumIncTreated <- calculateCumInc(treated)#a=1
cumIncPlacebo <- calculateCumInc(placebo)#a=0

gcompcderr<-cumIncTreated[length(cutTimes)]/cumIncPlacebo[length(cutTimes)] 
gcompcderd<-cumIncTreated[length(cutTimes)]-cumIncPlacebo[length(cutTimes)] 
print(gcompcderr)
print(gcompcderd)
```

## Total effect

We are going to use the same code as in the previous section, except that in this case instead of fixing the hazard of other causes to death as 0, we are going to calculate hazards for this outcome, fit from the `plrfit0` model and recalculate the survival probability, given death

For A = 1

```{r}
treated$hazardO <- predict(plrFitO, newdata = treated, type = 'response')

treated$s <- (1-treated$hazardP) * (1-treated$hazardO)

```

For A = 0

```{r}
placebo$hazardO <- predict(plrFitO, newdata = placebo, type = 'response')

placebo$s <- (1-placebo$hazardP) * (1-placebo$hazardO)
```

Calculate risks and corresponding treatment effects (RR/RD) by end of follow-up using parametric g-formula

```{r}
cumIncTreated <- calculateCumInc(treated)#a=1
cumIncPlacebo <- calculateCumInc(placebo)#a=0

gcomptotrr<-cumIncTreated[length(cutTimes)]/cumIncPlacebo[length(cutTimes)] 
gcomptotrd<-cumIncTreated[length(cutTimes)]-cumIncPlacebo[length(cutTimes)]
print(gcomptotrr)
print(gcomptotrd)
```

## Total effect of death on other causes

```{r}
# parametric g-formula estimate of the risk of the competing event itself
cumIncTreatedcomp <- calculateCumInc(treated, competing = TRUE)#a=1

cumIncPlacebocomp <- calculateCumInc(placebo, competing = TRUE)#a=0

gcompcomprr <-
  cumIncTreatedcomp[length(cutTimes)] / cumIncPlacebocomp[length(cutTimes)]

gcompcomprd <-
  cumIncTreatedcomp[length(cutTimes)] - cumIncPlacebocomp[length(cutTimes)]


print(gcompcomprr)
print(gcompcomprd)
```

## Separable effects

Following equation (9) from [Stensrud et al. Journal of the American Statistical Association. 2020](https://www.tandfonline.com/doi/full/10.1080/01621459.2020.1765783)

To this matter, we will use the long format data set prepared before: `longProstRed`. 

- Make copies of rx to be used in the hazard of other death

```{r}
longProstRed$Orx <- longProstRed$rx
```

- Fit conditional pooled logistic regression models

```{r}
plrFitP <-
    glm(
      prostateDeath ~ rx*(dtime + I(dtime ^ 2)+ I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
      data = longProstRed,
      family = binomial()
    )
    
plrFitO <-
  glm(
    otherDeath ~ Orx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
    data = longProstRed,
    family = binomial()
  )
```
  
- Create simulated data where everyone is treated 

Expand baseline so it contains a visit at each time point for every individual and where the baseline information has been carried forward at each time

```{r}
treated <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time

treated$dtime <- rep(cutTimes, n)
  
treated$rx <- 1

treated$Orx <- 1
  
```

- Make analogous dataset for placebo

```{r}
placebo <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time
  
placebo$dtime <- rep(cutTimes, n)

placebo$rx <- 0

placebo$Orx <- 0

```

Turn into conditional risk of event in each time interval

```{r}
treated$hazardP <-
    predict(plrFitP, newdata = treated, type = 'response')
  
treated$hazardO <-
    predict(plrFitO, newdata = treated, type = 'response')
  
treated$s <- (1 - treated$hazardP) * (1 - treated$hazardO)
```

```{r}
placebo$hazardP <-
    predict(plrFitP, newdata = placebo, type = 'response')

placebo$hazardO <-
    predict(plrFitO, newdata = placebo, type = 'response')

placebo$s <- (1 - placebo$hazardP) * (1 - placebo$hazardO)
```

- Make analogous dataset for Ay=1, Ad=0

```{r}
treatAy <-
    baseline[rep(1:n, each = length(cutTimes)), ]
  
treatAy$dtime <- rep(cutTimes, n)
  
treatAy$rx <- 1
  
treatAy$Orx <- 0

treatAy$hazardP <-
    predict(plrFitP, newdata = treatAy, type = 'response')
  
treatAy$hazardO <-
    predict(plrFitO, newdata = treatAy, type = 'response')
  
treatAy$s <- (1 - treatAy$hazardP) * (1 - treatAy$hazardO)

```

- Calculate cumulative incidence for each estimand

```{r}
cumIncTreated <- calculateCumInc(treated)
cumIncTreatAy <- calculateCumInc(treatAy)
cumIncPlacebo <- calculateCumInc(placebo)

estimatesFullData <-
  data.frame(cumIncTreated, cumIncTreatAy, cumIncPlacebo)

knitr::kable(round(estimatesFullData[36,], 2))
```

- Cumulative incidence curves

```{r}
plot(cutTimes, cumIncTreated, type = "s",
     ylim = c(0, 1), ylab = "Risk",
     xlab = "Months", 
     lty = 1, lwd = 1, xlim = c(0, 52))

lines(cutTimes, cumIncPlacebo, type = "s",
      col = 2, ylim = c(0, 1), lwd = 1)

lines(cutTimes, cumIncTreatAy, type = "s",
      col = 3, ylim = c(0, 1), lwd = 1)

legend(x = 0, y = 1, 
       c("(Ay=1,Ad=1)", "(Ay=1,Ad=0)", "(Ay=0,Ad=0)"),
       bty = "n", col = c(1, 3, 2),
       lty = c(1, 1, 1), cex = 1, pt.cex = 1, lwd = 1)
```


### [Go to: Data cleaning and setup here](https://palolili23.github.io/2022_ser_competing_events/R/index.html)

### [Go to: Direct, total and separable effects, using IPW](https://palolili23.github.io/2022_ser_competing_events/R/ipw.html)