---
title: "Causal inference with competing events"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Data cleaning and setup

### Initial steps

- Load the required packages

```{r}
library(Hmisc)
library(splines)
library(here)
```

- Download the data

The example dataset is Byar & Greene prostate cancer data, downloaded from the following [website](https://hbiostat.org/data/). The data dictionary available [here](https://hbiostat.org/data/repo/Cprostate.html).
**Note**: Please note that there are many slightly different versions of this dataset, and the one in this website is not identical to the one used in Young et al. and Stensrud et al. For this workshop, we recommend using the version of this data that is located in this repository. [Link](https://github.com/palolili23/2022_ser_competing_events/blob/main/R/prostate.csv)

```{r}
prostate <- read.csv(here::here("R", "prostate.csv"), sep = ';')
```

- Load utility functions, these are located in the github repository. These functions are necessary to calculate hazards and cumulative incidence. [Link](https://github.com/palolili23/2022_ser_competing_events/blob/main/R/utility_functions.R)

```{r}
source(here::here("R", "utility_functions.R"))
```

### Data cleaning

The goal is to create a data set in long format, so that every participant has as many rows as time-points of follow-up. New variables will be created as indicators of time, outcome, competing event and censoring event.


- Create a binary variable to indicate all-cause death

```{r}
prostate$allCause <- prostate$status != "alive"
```

- Create a variable with 3 levels indicating the cause of death 

```{r}
prostate$eventType <- as.factor(prostate$status)

levels(prostate$eventType) <-
  list(alive = "alive",
       pdeath = "dead - prostatic ca",
       odeath = c(levels(prostate$eventType)[c(2:5, 7:10)]))
```

- Filter data to only include high dose DES (A=1) and placebo (A=0), clean additional covariates

```{r}
prostate$rx <- as.factor(prostate$rx)
prostRed <- prostate[prostate$rx %in% levels(prostate$rx)[3:4], ]
prostRed$temprx = as.integer(prostRed$rx) - 3
prostRed$rx = abs(1 - prostRed$temprx) # DES is A=1 and placebo A=0
prostRed$eventType = as.integer(prostRed$eventType) - 1 #0 is censoring, 1 is pdeath, 2 is odeath
prostRed$hgBinary <- prostRed$hg < 12
prostRed$ageCat <- cut2(prostRed$age, c(0, 60, 70, 80, 100))
prostRed$normalAct <- prostRed$pf == "normal activity"
prostRed$eventCens <- prostRed$eventType == 0
prostRed$eventProst <- prostRed$eventType == 1
```

- Descriptive information

```{r}
table(prostRed$eventType, prostRed$rx)
```

### Convert data to long format

```{r}
prostRed$Tstart = -0.01 #Needed for the long format

cutTimes <- c(0:59)

longProstRed <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "allCause")

longProstRedCens <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "eventCens")

longProstRed$eventCens <- longProstRedCens$eventCens

# Make column for prostate cancer mortality
longProstRed$prostateDeath <-
  longProstRed$allCause == 1 &  longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2

```

- Glimpse of data structure, for this matter we select 2 participants with different outcomes.

```{r,  echo = FALSE}
longProstRed %>% 
  filter(patno %in% c(462, 480)) %>% select(
    patno, dtime, eventCens, allCause, prostateDeath, otherDeath
  ) %>% knitr::kable()
```

- To be sure that right risk sets are used in models for this data structure without explicit conditioning, set "future values" of the competing event (D) and the outcome of interest (Y) to missing on a given line if C = 1 on that line and set future values of Y to missing on a line if D = 1 on that line (respects "order" C, D, Y in each interval). This may not be of an issue when there are no "ties" but with month long or bigger intervals intervals there might be.

```{r}
longProstRed$prostateDeath[longProstRed$eventCens == 1] <- NA

longProstRed$otherDeath[longProstRed$eventCens == 1] <- NA

longProstRed$prostateDeath[longProstRed$otherDeath == 1] <- NA
```


```{r, echo = FALSE}
longProstRed %>% 
  filter(patno %in% c(462, 480)) %>% select(
    patno, dtime, eventCens, allCause, prostateDeath, otherDeath
  ) %>% knitr::kable()
```

- Restrict input data set to records with dtime < K + 1 for fitting pooled over time models

```{r}

# Before
dim(longProstRed)

longProstRed <- longProstRed[longProstRed$dtime < length(cutTimes), ]

# After filtering records 

dim(longProstRed)

```

- Create 'baseline' - data collected at visit 0

```{r}
baseline <- longProstRed[longProstRed$dtime == 0, ]

dim(baseline)
```

- Save number of subjects

```{r}
n <- length(unique(longProstRed$patno))

n
```


# [Go to: Direct, total and separable effects, using IPW](https://palolili23.github.io/2022_ser_competing_events/R/ipw.html)

# [Go to: Direct, total and separable effects, using G-formula](https://palolili23.github.io/2022_ser_competing_events/R/gform.html)

# References

- Young JG, Stensrud MJ, Tchetgen Tchetgen EJ, Hernán MA. A causal framework for classical statistical estimands in failure time settings with competing events. Statistics in Medicine 2020. [Paper link](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471) [Code](https://github.com/CausalInference/CompetingEvents_Young_SIM_2020)

- Stensrud MJ, Young JG, Didelez V, Robins JM, Hernán MA. Separable Effects for Causal Inference in the Presence of Competing Events. Journal of the American Statistical Association 2022. [Paper link](https://www.tandfonline.com/doi/full/10.1080/01621459.2020.1765783)



