---
title: "Causal inference with competing events"
output: 
  html_document:
    date: "`r Sys.Date()`"
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    theme: flatly
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
```

## Setup

- Load the required packages

```{r, }
library(Hmisc)
library(splines)
library(here)
```

- Download the data

The example dataset is Byar & Greene prostate cancer data, downloaded from the following [website](https://hbiostat.org/data/). Data dictionary available [here](https://hbiostat.org/data/repo/Cprostate.html)

```{r}
prostate <- read.csv(here::here("R", "prostate.csv"), sep = ';')
```

- Load utility functions

```{r}
source(here::here("R", "utility_functions.R"))
```

### Data cleaning

- Create a binary variable to indicate all-cause death

```{r}
prostate$allCause <- prostate$status != "alive"
```

- Create a variable with 3 levels indicating the cause of death 

```{r}
# prostate$eventType <- prostate$status
# 
# levels(prostate$eventType) <- list(alive="alive",pdeath="dead - prostatic ca",odeath = 
#                                      c(levels(prostate$eventType)[c(2:5,7:10)]))
prostate <- prostate %>% 
  mutate(eventType = case_when(status == "alive" ~ 0,
                      status == "dead - prostatic ca" ~ 1,
                      TRUE ~ 2))
```

- Filter data to only include high dose DES (A=1) and placebo (A=0)

```{r}
prostate$rx <- as.factor(prostate$rx)
prostRed <- prostate[prostate$rx %in% levels(prostate$rx)[3:4],]
prostRed$temprx = as.integer(prostRed$rx)-3 
prostRed$rx=abs(1-prostRed$temprx) ## DES is A=1 and placebo A=0
# prostRed$eventType = as.integer(prostRed$eventType)-1 #0 is censoring, 1 is pdeath, 2 is odeath
prostRed$hgBinary <- prostRed$hg < 12
prostRed$ageCat <- cut2(prostRed$age,c(0,60,70,80,100))
prostRed$normalAct <- prostRed$pf == "normal activity"
prostRed$eventCens <- prostRed$eventType == 0
prostRed$eventProst <- prostRed$eventType == 1
prostRed$Tstart = -0.01 #Needed for the long format

```

- Descriptive information

```{r}
table(prostRed$eventType, prostRed$rx)
```

### Convert data to long format

```{r}
cutTimes <- c(0:59)

longProstRed <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "allCause"
  )

longProstRedCens <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "eventCens"
  )

longProstRed$eventCens <- longProstRedCens$eventCens

# Make column for prostate cancer mortality
longProstRed$prostateDeath <-
  longProstRed$allCause == 1 &  longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2

```

To be sure that right risk sets are used in models for this data structure without explicit conditioning, set "future values" of D and Y to missing on a given line if C=1 on that line and set future values of Y to missing on a line if D=1 on that line (respects "order" C,D,Y in each interval) -- nonissue when there are no "ties" but with month long intervals there might be.

```{r}
longProstRed$prostateDeath[longProstRed$eventCens == 1] <- NA

longProstRed$otherDeath[longProstRed$eventCens == 1] <- NA

longProstRed$prostateDeath[longProstRed$otherDeath == 1] <- NA
```

Restrict input data set to records with dtime<K+1 for fitting pooled over time models

```{r}
longProstRed <- longProstRed[longProstRed$dtime < length(cutTimes), ]
```

Create 'baseline' - data collected at visit 0

```{r}
baseline <- longProstRed[longProstRed$dtime == 0, ]
```

Save number of subjects

```{r}
n <- length(unique(longProstRed$patno))
```

### Construction of weights

- Fit conditional pooled logistic regression models for:

```{r}
plrFitP <-
  glm(
    prostateDeath ~ rx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary ,
    data = longProstRed,
    family = binomial()
  )

plrFitO <-
  glm(
    otherDeath ~ dtime + I(dtime ^ 2) + normalAct + ageCat + hx + hgBinary + rx ,
    data = longProstRed,
    family = binomial()
  )

temp50 <- longProstRed[longProstRed$dtime > 50,]

plrFitC <-
  glm(eventCens ~  normalAct + ageCat + hx + rx,
      data = temp50,
      family = binomial())

```

- Fit models for the numerators in the stabilized weights

```{r}
plrFitCnum <- glm(eventCens ~ rx , data = temp50, family = binomial())

plrFitOnum <-
  glm(otherDeath ~ dtime + I(dtime ^ 2) + rx,
      data = longProstRed,
      family = binomial())

```

- Create unstabilized and stabilized weights for censoring due to loss to follow-up

```{r}
### Denominator

longProstRed$predC <-
  rep(1, dim(longProstRed)[1]) # start by setting all to 1

longProstRed$predC[longProstRed$dtime > 50] <-
  1 - predict(plrFitC, newdata = temp50, type = 'response') #replace with model-based values for records with dtime>50

### Numerator

longProstRed$predCnum <-
  rep(1, dim(longProstRed)[1])

longProstRed$predCnum [longProstRed$dtime > 50] <-
  1 - predict(plrFitCnum, newdata = temp50, type = 'response') #replace with model-based values for records with dtime>50

predC <- longProstRed$predC

predCnum <- longProstRed$predCnum

cumPredC <-
  unlist(aggregate(predC ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F) #unlist because aggregate creates list

cumPredCnum <-
  unlist(aggregate(predCnum ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F)

### Unstabilized weights
weights_te <- 1 / cumPredC

### Stabilized weights

weights_te_stab <- ipwEstimand_te_stab <- cumPredCnum / cumPredC
```

- Create unstabilized and stabilized weights for death as a censoring event

```{r}

### Denominator

predO <-
  1 - predict(plrFitO, newdata = longProstRed, type = 'response')

cumPredO <-
  unlist(aggregate(predO ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F) #unlist because aggregate creates list
### Numerator

predOnum <-
  1 - predict(plrFitOnum, newdata = longProstRed, type = 'response')

cumPredOnum <-
  unlist(aggregate(predOnum ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F) 

weights_de <- 1 / cumPredC * 1 / cumPredO

weights_de_stab <-
  cumPredCnum / cumPredC * cumPredOnum / cumPredO
```

- Assess weights

```{r}
summary(weights_te)

summary(weights_de)

summary(weights_te_stab)

summary(weights_de_stab)
```

## Direct effect

```{r}

### Calculate cumulative hazards and cumulative incidence for the treated arm

#### Unstabilized
Treated_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE
  )

Treated_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_de <-
  nonParametricCumInc(Treated_de_hazardP, Treated_de_hazardO)

#### Stabilized
Treated_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE
  )

Treated_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_des <-
  nonParametricCumInc(Treated_de_hazardPs, Treated_de_hazardOs)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

#### Unstabilized

Placebo_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_de <-
  nonParametricCumInc(Placebo_de_hazardP, Placebo_de_hazardO)

#### Stabilized

Placebo_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_des <-
  nonParametricCumInc(Placebo_de_hazardPs, Placebo_de_hazardOs)

### Calculate risk difference and risk ratio

de_rr <-
  cumIncTreated_de[length(cutTimes)] / cumIncPlacebo_de[length(cutTimes)]

de_rd <-
  cumIncTreated_de[length(cutTimes)] - cumIncPlacebo_de[length(cutTimes)]

print(de_rr)

print(de_rd)

```

## Total effect

```{r}

### Calculate cumulative hazards and cumulative incidence for the treated arm

#### Unstabilized

Treated_te_hazardP <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE
  )

Treated_te_hazardO <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = FALSE
  )

cumIncTreated_te <-
  nonParametricCumInc(Treated_te_hazardP, Treated_te_hazardO)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

#### Unstabilized

Placebo_te_hazardP <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_te_hazardO <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = FALSE
  )

cumIncPlacebo_te <-
  nonParametricCumInc(Placebo_te_hazardP, Placebo_te_hazardO)

### Calculate risk difference and risk ratio

te_rr <-
  cumIncTreated_te[length(cutTimes)] / cumIncPlacebo_te[length(cutTimes)]

te_rd <-
  cumIncTreated_te[length(cutTimes)] - cumIncPlacebo_te[length(cutTimes)]

print(te_rr)
print(te_rd)
```

## Total effect on death for other causes

```{r}
cumIncTreated_te_comp <-
  nonParametricCumInc(Treated_te_hazardO, Treated_te_hazardP, competing = TRUE)

cumIncPlacebo_te_comp <-
  nonParametricCumInc(Placebo_te_hazardO, Placebo_te_hazardP, competing = TRUE)

te_comp_rr <-
  cumIncTreated_te_comp[length(cutTimes)] / cumIncPlacebo_te_comp[length(cutTimes)]

te_comp_rd <-
  cumIncTreated_te_comp[length(cutTimes)] - cumIncPlacebo_te_comp[length(cutTimes)]

print(te_comp_rr)

print(te_comp_rd)
```

## Separable effects

- Convert data to long format, up to 100 months

```{r}
cutTimes <- c(0:100)

longProstRed <-
    survSplit(
      data = prostRed,
      cut = cutTimes,
      start = "Tstart",
      end = "dtime",
      event = "allCause"
    )

longProstRedCens <-
    survSplit(
      data = prostRed,
      cut = cutTimes,
      start = "Tstart",
      end = "dtime",
      event = "eventCens"
    )

longProstRed$eventCens <- longProstRedCens$eventCens
```

- Make column for prostate cancer mortality

```{r}
longProstRed$prostateDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2
```

```{r}
n <- length(unique(longProstRed$patno))
```
  
- Create 'baseline' - data collected at visit 0

```{r}
baseline <- longProstRed[longProstRed$dtime == 0,]
```

- Make copies of rx to be used in the hazard of other death

```{r}
longProstRed$Orx <- longProstRed$rx
```

```{r}
plrFitP <-
    glm(
      prostateDeath ~ rx*(dtime + I(dtime ^ 2)+ I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
      data = longProstRed,
      family = binomial()
    )
    
plrFitO <-
  glm(
    otherDeath ~ Orx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
    data = longProstRed,
    family = binomial()
  )
```
  
- Create simulated data where everyone is treated 

Expand baseline so it contains a visit at each time point for every individual and where the baseline information has been carried forward at each time

```{r}
treated <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time

treated$dtime <- rep(cutTimes, n)
  
treated$rx <- 0

treated$Orx <- 0
  
treated$hazardP <-
    predict(plrFitP, newdata = treated, type = 'response')
  
treated$hazardO <-
    predict(plrFitO, newdata = treated, type = 'response')
  
treated$s <- (1 - treated$hazardP) * (1 - treated$hazardO)

sum(treated$hazardO < 0)
```


- Make analogous dataset for placebo

```{r}
placebo <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time
  
placebo$dtime <- rep(cutTimes, n)

placebo$rx <- 1

placebo$Orx <- 1

# Turn into conditional risk of event in each time interval
placebo$hazardP <-
    predict(plrFitP, newdata = placebo, type = 'response')

placebo$hazardO <-
    predict(plrFitO, newdata = placebo, type = 'response')

placebo$s <- (1 - placebo$hazardP) * (1 - placebo$hazardO)
  
```

- Make analogous dataset for Ay=0, Ad=1

```{r}
treatAy <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time
  
treatAy$dtime <- rep(cutTimes, n)
  
treatAy$rx <- 0
  
treatAy$Orx <- 1

# Turn into conditional risk of event in each time interval

treatAy$hazardP <-
    predict(plrFitP, newdata = treatAy, type = 'response')
  
treatAy$hazardO <-
    predict(plrFitO, newdata = treatAy, type = 'response')
  
treatAy$s <- (1 - treatAy$hazardP) * (1 - treatAy$hazardO)

```

- Calculate cumulative incidence for each estimand

```{r}
cumIncTreated <- calculateCumInc(treated)
cumIncTreatAy <- calculateCumInc(treatAy)
cumIncPlacebo <- calculateCumInc(placebo)

estimatesFullData <-
  data.frame(cumIncTreated, cumIncTreatAy, cumIncPlacebo)
```

- Cumulative incidence curves

```{r}
riskTreated <- estimatesFullData$cumIncTreated
riskTreatAy <- estimatesFullData$cumIncTreatAy
riskPlacebo <- estimatesFullData$cumIncPlacebo

plot(cutTimes, riskTreated, type = "s",
     ylim = c(0, 1), ylab = "Cumulative incidence",
     xlab = "Months", 
     lty = 1, lwd = 1, xlim = c(0, 52))

lines(cutTimes, riskPlacebo, type = "s",
      col = 2, ylim = c(0, 1), lwd = 1)

lines(cutTimes, riskTreatAy, type = "s",
      col = 3, ylim = c(0, 1), lwd = 1)

legend(x = 0, y = 1, 
       c("(Ay=1,Ad=1)", "(Ay=1,Ad=0)", "(Ay=0,Ad=0)"),
       bty = "n", col = c(1, 3, 2),
       lty = c(1, 1, 1), cex = 1, pt.cex = 1, lwd = 1)
```

## References

- Young JG, Stensrud MJ, Tchetgen Tchetgen EJ, Hernán MA. A causal framework for classical statistical estimands in failure time settings with competing events. Statistics in Medicine 2020. [Paper link](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471) [Code](https://github.com/CausalInference/CompetingEvents_Young_SIM_2020)

- Stensrud MJ, Young JG, Didelez V, Robins JM, Hernán MA. Separable Effects for Causal Inference in the Presence of Competing Events. Journal of the American Statistical Association 2022. [Paper link](https://www.tandfonline.com/doi/full/10.1080/01621459.2020.1765783)



