---
title: "ipw"
output: html_document
date: "2023-05-31"
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Hmisc)
library(splines)
library(here)
source(here::here("R", "utility_functions.R"))

prostate <- read.csv(here::here("R", "prostate.csv"), sep = ';')
prostate$allCause <- prostate$status != "alive"
prostate$eventType <- as.factor(prostate$status)

levels(prostate$eventType) <-
  list(alive = "alive",
       pdeath = "dead - prostatic ca",
       odeath = c(levels(prostate$eventType)[c(2:5, 7:10)]))

prostate$rx <- as.factor(prostate$rx)
prostRed <- prostate[prostate$rx %in% levels(prostate$rx)[3:4], ]
prostRed$temprx = as.integer(prostRed$rx) - 3
prostRed$rx = abs(1 - prostRed$temprx) # DES is A=1 and placebo A=0
prostRed$eventType = as.integer(prostRed$eventType) - 1 #0 is censoring, 1 is pdeath, 2 is odeath
prostRed$hgBinary <- prostRed$hg < 12
prostRed$ageCat <- cut2(prostRed$age, c(0, 60, 70, 80, 100))
prostRed$normalAct <- prostRed$pf == "normal activity"
prostRed$eventCens <- prostRed$eventType == 0
prostRed$eventProst <- prostRed$eventType == 1
prostRed$Tstart = -0.01 #Needed for the long format
cutTimes <- c(0:59)

longProstRed <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "allCause")

longProstRedCens <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "eventCens")

longProstRed$eventCens <- longProstRedCens$eventCens

longProstRed$prostateDeath <-
  longProstRed$allCause == 1 &  longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2

longProstRed$prostateDeath[longProstRed$eventCens == 1] <- NA

longProstRed$otherDeath[longProstRed$eventCens == 1] <- NA

longProstRed$prostateDeath[longProstRed$otherDeath == 1] <- NA

longProstRed <- longProstRed[longProstRed$dtime < length(cutTimes), ]

baseline <- longProstRed[longProstRed$dtime == 0, ]

n <- length(unique(longProstRed$patno))

```


### Computation of IPC weights

- First, we run a set of conditional pooled logistic models, both as a first step to calculate the risk of prostate cancer death, and to calculate weights for death due to other causes, and for censoring.  

```{r}
### Pooled logistic regression for prostate cancer death (this will be needed in the next section)
plrFitP <-
  glm(
    prostateDeath ~ rx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary ,
    data = longProstRed,
    family = binomial())

### Pooled logistic regression for death due to other causes

plrFitO <-
  glm(
    otherDeath ~ dtime + I(dtime ^ 2) + normalAct + ageCat + hx + hgBinary + rx ,
    data = longProstRed,
    family = binomial())


### Pooled logistic regression for censoring after time 50

temp50 <- longProstRed[longProstRed$dtime > 50,]

plrFitC <-
  glm(eventCens ~  normalAct + ageCat + hx + rx,
      data = temp50,
      family = binomial())

```

- Fit models for the numerators in the stabilized weights

```{r}
plrFitCnum <-
  glm(eventCens ~ rx , data = temp50, family = binomial())

plrFitOnum <-
  glm(otherDeath ~ dtime + I(dtime ^ 2) + rx,
      data = longProstRed,
      family = binomial())

```

- Calculate weights for censoring due to loss to follow-up

```{r}
### Denominator

longProstRed$predC <-
  rep(1, dim(longProstRed)[1]) # The probability of *not* being censored is 1 for all t <= 50

longProstRed$predC[longProstRed$dtime > 50] <- # update this probability from time 50 and onwards
  1 - predict(plrFitC, newdata = temp50, type = 'response')

### Numerator

longProstRed$predCnum <-
  rep(1, dim(longProstRed)[1])

longProstRed$predCnum [longProstRed$dtime > 50] <-
  1 - predict(plrFitCnum, newdata = temp50, type = 'response') 

predC <- longProstRed$predC

predCnum <- longProstRed$predCnum

cumPredC <-
  unlist(aggregate(predC ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F)

cumPredCnum <-
  unlist(aggregate(predCnum ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F)

### Unstabilized weights
weights_te <- 1 / cumPredC

### Stabilized weights

weights_te_stab <- ipwEstimand_te_stab <- cumPredCnum / cumPredC
```

- Calculate unstabilized and stabilized weights for death from other causes

```{r}

### Denominator

predO <-
  1 - predict(plrFitO, newdata = longProstRed, type = 'response')

cumPredO <-
  unlist(aggregate(predO ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F)

### Numerator

predOnum <-
  1 - predict(plrFitOnum, newdata = longProstRed, type = 'response')

cumPredOnum <-
  unlist(aggregate(predOnum ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F) 

weights_de <- 1 / cumPredC * 1 / cumPredO

weights_de_stab <-
  cumPredCnum / cumPredC * cumPredOnum / cumPredO
```

- Assess weights

```{r}
summary(weights_te)

summary(weights_de)

summary(weights_te_stab)

summary(weights_de_stab)
```

## Direct effect

Following equation (37) from [Young et al. Statistics in Medicine. 2020](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471).

```{r}

### Calculate cumulative hazards and cumulative incidence for the treated arm

#### Unstabilized
Treated_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_de <-
  nonParametricCumInc(Treated_de_hazardP, Treated_de_hazardO)

#### Stabilized
Treated_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_des <-
  nonParametricCumInc(Treated_de_hazardPs, Treated_de_hazardOs)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

#### Unstabilized

Placebo_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_de <-
  nonParametricCumInc(Placebo_de_hazardP, Placebo_de_hazardO)

#### Stabilized

Placebo_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_des <-
  nonParametricCumInc(Placebo_de_hazardPs, Placebo_de_hazardOs)

```

- Risks at 60 months of follow-up, risk difference and risk ratio


```{r}
cumIncTreated_de_60 <- cumIncTreated_de[length(cutTimes)] 

cumIncPlacebo_de_60 <- cumIncPlacebo_de[length(cutTimes)]

de_rr <- cumIncTreated_de_60 / cumIncPlacebo_de_60

de_rd <- cumIncTreated_de_60 - cumIncPlacebo_de_60

results_de <- data.frame(cumIncTreated_de_60, cumIncPlacebo_de_60, de_rr, de_rd)

knitr::kable(round(results_de, 2)) 
```

- Cumulative incidence curves

```{r}
plot(
  cutTimes, cumIncTreated_de, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of prostate cancer death"),
    paste("under elimination of competing events")
  )
)
lines(
  cutTimes,
  cumIncPlacebo_de,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2),
  lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)
```

## Total effect

Following equation (40) from [Young et al. Statistics in Medicine. 2020](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471).

```{r}

### Calculate cumulative hazards and cumulative incidence for the treated arm

#### Unstabilized

Treated_te_hazardP <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_te_hazardO <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = FALSE)

cumIncTreated_te <-
  nonParametricCumInc(Treated_te_hazardP, Treated_te_hazardO)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

#### Unstabilized

Placebo_te_hazardP <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE)

Placebo_te_hazardO <-
  nonParametricCumHaz(
    weights_te,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = FALSE)

cumIncPlacebo_te <-
  nonParametricCumInc(Placebo_te_hazardP, Placebo_te_hazardO)

```

- Risks at 60 months of follow-up, risk difference and risk ratio


```{r}
cumIncTreated_te_60 <- cumIncTreated_te[length(cutTimes)] 

cumIncPlacebo_te_60 <- cumIncPlacebo_te[length(cutTimes)]

te_rr <- cumIncTreated_te_60 / cumIncPlacebo_te_60

te_rd <- cumIncTreated_te_60 - cumIncPlacebo_te_60

results_te <- data.frame(cumIncTreated_te_60, cumIncPlacebo_te_60, te_rr, te_rd)

knitr::kable(round(results_te, 2))
```

- Cumulative incidence curves

```{r}
plot(
  cutTimes, cumIncTreated_te, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of prostate cancer death"),
    paste("without elimination of competing events")
  )
)
lines(
  cutTimes,
  cumIncPlacebo_te,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2), lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)

```

## Total effect on death for other causes

```{r}
cumIncTreated_te_comp <-
  nonParametricCumInc(Treated_te_hazardO, Treated_te_hazardP, competing = TRUE)

cumIncPlacebo_te_comp <-
  nonParametricCumInc(Placebo_te_hazardO, Placebo_te_hazardP, competing = TRUE)

```

```{r}
cumIncTreated_te_comp_60 <- cumIncTreated_te_comp[length(cutTimes)] 

cumIncPlacebo_te_comp_60 <- cumIncPlacebo_te_comp[length(cutTimes)]

te_comp_rr <- cumIncTreated_te_comp_60 / cumIncPlacebo_te_comp_60

te_comp_rd <- cumIncTreated_te_comp_60 - cumIncPlacebo_te_comp_60

results_te_comp <- data.frame(cumIncTreated_te_comp_60, cumIncPlacebo_te_comp_60, te_comp_rr, te_comp_rd)

knitr::kable(round(results_te_comp, 2))

```

- Cumulative incidence curves

```{r}
plot(
  cutTimes,
  cumIncTreated_te_comp, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of other causes of death (the competing event)")
  ))

lines(
  cutTimes,
  cumIncPlacebo_te_comp,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2), lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)

```

## Separable effects