---
title: "Direct, total and separable effects, using inverse probability weighting (IPW)"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

```{r, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Hmisc)
library(splines)
library(here)
source(here::here("R", "utility_functions.R"))

prostate <- read.csv(here::here("R", "prostate_mats.csv"))
prostate$allCause <- prostate$status != "alive"
prostate$eventType <- as.factor(prostate$status)

levels(prostate$eventType) <-
  list(alive = "alive",
       pdeath = "dead - prostatic ca",
       odeath = c(levels(prostate$eventType)[c(2:5, 7:10)]))

prostate$rx <- as.factor(prostate$rx)
prostRed <- prostate[prostate$rx %in% levels(prostate$rx)[3:4], ]
prostRed$temprx = as.integer(prostRed$rx) - 3
prostRed$rx = abs(1 - prostRed$temprx) # DES is A=1 and placebo A=0
prostRed$eventType = as.integer(prostRed$eventType) - 1 #0 is censoring, 1 is pdeath, 2 is odeath
prostRed$hgBinary <- prostRed$hg < 12
prostRed$ageCat <- cut2(prostRed$age, c(0, 60, 70, 80, 100))
prostRed$normalAct <- prostRed$pf == "normal activity"
prostRed$eventCens <- prostRed$eventType == 0
prostRed$eventProst <- prostRed$eventType == 1
prostRed$Tstart = -0.01 #Needed for the long format
cutTimes <- c(0:59)

longProstRed <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "allCause")

longProstRedCens <-
  survSplit(
    data = prostRed,
    cut = cutTimes,
    start = "Tstart",
    end = "dtime",
    event = "eventCens")

longProstRed$eventCens <- longProstRedCens$eventCens

longProstRed$prostateDeath <-
  longProstRed$allCause == 1 &  longProstRed$eventType == 1

longProstRed$otherDeath <-
  longProstRed$allCause == 1 & longProstRed$eventType == 2

longProstRed$prostateDeath[longProstRed$eventCens == 1] <- NA

longProstRed$otherDeath[longProstRed$eventCens == 1] <- NA

longProstRed$prostateDeath[longProstRed$otherDeath == 1] <- NA

longProstRed <- longProstRed[longProstRed$dtime < length(cutTimes), ]

baseline <- longProstRed[longProstRed$dtime == 0, ]

n <- length(unique(longProstRed$patno))

```


## Conditional pooled logistic models

- First, we run a set of conditional pooled logistic models, both as a first step to calculate the risk of prostate cancer death, and to calculate weights for death due to other causes, and weights for censoring due to loss to follow-up

```{r}
### Pooled logistic regression for prostate cancer death (this will be needed in the next section)
plrFitP <-
  glm(
    prostateDeath ~ rx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary ,
    data = longProstRed,
    family = binomial())

### Pooled logistic regression for death due to other causes

plrFitO <-
  glm(
    otherDeath ~ dtime + I(dtime ^ 2) + normalAct + ageCat + hx + hgBinary + rx ,
    data = longProstRed,
    family = binomial())


### Pooled logistic regression for censoring after time 50

temp50 <- longProstRed[longProstRed$dtime > 50,]

plrFitC <-
  glm(eventCens ~  normalAct + ageCat + hx + rx,
      data = temp50,
      family = binomial())

```

- Fit models for the numerators in the stabilized weights corresponding to death due to other causes, and for censoring

```{r}
plrFitCnum <-
  glm(eventCens ~ rx , data = temp50, family = binomial())

plrFitOnum <-
  glm(otherDeath ~ dtime + I(dtime ^ 2) + rx,
      data = longProstRed,
      family = binomial())

```

## Weights calculation

### Calculate weights for censoring due to loss to follow-up

```{r}
### Denominator

longProstRed$predC <-
  rep(1, dim(longProstRed)[1]) # The probability of *not* being censored is 1 for all t <= 50

longProstRed$predC[longProstRed$dtime > 50] <- # update this probability from time 50 and onwards
  1 - predict(plrFitC, newdata = temp50, type = 'response')

### Numerator

longProstRed$predCnum <-
  rep(1, dim(longProstRed)[1])

longProstRed$predCnum [longProstRed$dtime > 50] <-
  1 - predict(plrFitCnum, newdata = temp50, type = 'response') 

predC <- longProstRed$predC

predCnum <- longProstRed$predCnum
```

- Glimpse of data:

```{r, echo = FALSE}
longProstRed %>% 
  filter(patno == 8) %>% slice(45:60) %>% 
  select(patno, dtime, starts_with("pred")) %>% 
  knitr::kable()
```

- Calculate the cumulative probability over time:

```{r}
cumPredC <-
  unlist(aggregate(predC ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F)

cumPredCnum <-
  unlist(aggregate(predCnum ~ longProstRed$patno, FUN = cumprod)$predC,
         use.names = F)

```

- Glimpse of data:

```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredC = cumPredC) %>% 
    filter(patno == 8) %>% slice(45:60) %>% 
  select(patno, dtime, starts_with("pred"), cumPredC) %>% 
  knitr::kable()
```

- Calculate unstabilized and stabilized weights

```{r}

### Unstabilized weights
weights_cens <- 1 / cumPredC

### Stabilized weights

weights_cens_stab <- cumPredCnum / cumPredC
```

```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredC = cumPredC,
         weights_cens = weights_cens,
         weights_cens_stab = weights_cens_stab) %>% 
    filter(patno == 8) %>% slice(45:60) %>% 
  select(patno, dtime, starts_with("pred"), cumPredC, starts_with("weig")) %>% 
  knitr::kable()
```


```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredC = cumPredC,
         weights_cens = weights_cens,
         weights_cens_stab = weights_cens_stab) %>% 
ggplot(aes(as_factor(dtime), weights_cens)) +
  geom_boxplot() +
  labs(title = "Unstabilized weights for censoring")
```


```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredC = cumPredC,
         weights_cens = weights_cens,
         weights_cens_stab = weights_cens_stab) %>% 
ggplot(aes(as_factor(dtime), weights_cens_stab)) +
  geom_boxplot() +
  labs(title = "Unstabilized weights for censoring")
```

### Calculate unstabilized and stabilized weights for death from other causes

```{r}
### Denominator

predO <-
  1 - predict(plrFitO, newdata = longProstRed, type = 'response')

### Numerator
predOnum <-
  1 - predict(plrFitOnum, newdata = longProstRed, type = 'response')

### Calculate cumulative probability

cumPredO <-
  unlist(aggregate(predO ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F)

cumPredOnum <-
  unlist(aggregate(predOnum ~ longProstRed$patno, FUN = cumprod)$predO,
         use.names = F) 

### Weight calculation

weights_compev <- 1 / cumPredO

weights_compev_stab <-
  cumPredOnum / cumPredO
```

- Glimpse of data

```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredO = cumPredO,
         cumPredOnum = cumPredOnum,
         weights_compev = weights_compev,
         weights_compev_stab = weights_compev_stab) %>% 
    filter(patno == 487) %>% 
  select(patno, dtime, cumPredO, cumPredOnum, starts_with("weig")) %>% 
  knitr::kable()

```

```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredO = cumPredO,
         cumPredOnum = cumPredOnum,
         weights_compev = weights_compev,
         weights_compev_stab = weights_compev_stab) %>% 
ggplot(aes(as_factor(dtime), weights_compev)) +
  geom_boxplot() +
  labs(title = "Unstabilized weights for death of other causes")
```

```{r, echo = FALSE}
longProstRed %>% 
  mutate(cumPredO = cumPredO,
         cumPredOnum = cumPredOnum,
         weights_compev = weights_compev,
         weights_compev_stab = weights_compev_stab) %>% 
ggplot(aes(as_factor(dtime), weights_compev_stab)) +
  geom_boxplot() +
  labs(title = "Stabilized weights for death of other causes")
```

- Multiply weights for censoring and for death due to other causes

```{r}
weights_de <- weights_cens * weights_compev

weights_de_stab <- weights_cens_stab * weights_compev_stab
```

- Assess weights

```{r}
summary(weights_de)

summary(weights_de_stab)
```

## Direct effect estimation

In this section we will apply utility functions that are located in the github repository. These functions are necessary to calculate hazards and cumulative incidence. [Link](https://github.com/palolili23/SER_competing_events_workshop/blob/main/R/utility_functions.R)

Following equation (37) from [Young et al. Statistics in Medicine. 2020](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471).

```{r}

### Calculate cumulative hazards and cumulative incidence for the treated arm

#### Unstabilized
Treated_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_de <-
  nonParametricCumInc(Treated_de_hazardP, Treated_de_hazardO)

#### Stabilized
Treated_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncTreated_des <-
  nonParametricCumInc(Treated_de_hazardPs, Treated_de_hazardOs)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

#### Unstabilized

Placebo_de_hazardP <-
  nonParametricCumHaz(
    weights_de,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardO <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_de <-
  nonParametricCumInc(Placebo_de_hazardP, Placebo_de_hazardO)

#### Stabilized

Placebo_de_hazardPs <-
  nonParametricCumHaz(
    weights_de_stab,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE
  )

Placebo_de_hazardOs <- rep(0, length.out = (length(cutTimes)))

cumIncPlacebo_des <-
  nonParametricCumInc(Placebo_de_hazardPs, Placebo_de_hazardOs)

```

- Risks under each treatment arm at 60 months of follow-up, risk difference and risk ratio

```{r}
## Unstabilized
cumIncTreated_de_60 <- cumIncTreated_de[length(cutTimes)] 

cumIncPlacebo_de_60 <- cumIncPlacebo_de[length(cutTimes)]

de_rr <- cumIncTreated_de_60 / cumIncPlacebo_de_60

de_rd <- cumIncTreated_de_60 - cumIncPlacebo_de_60

results_de <- data.frame(cumIncTreated_de_60, cumIncPlacebo_de_60, de_rr, de_rd)

knitr::kable(round(results_de, 2)) 
```

```{r}
## Stabilized

cumIncTreated_des_60 <- cumIncTreated_des[length(cutTimes)] 

cumIncPlacebo_des_60 <- cumIncPlacebo_des[length(cutTimes)]

des_rr <- cumIncTreated_des_60 / cumIncPlacebo_des_60

des_rd <- cumIncTreated_des_60 - cumIncPlacebo_des_60

results_des <- data.frame(cumIncTreated_des_60, cumIncPlacebo_des_60, des_rr, des_rd)

knitr::kable(round(results_des, 2)) 
```

- Cumulative incidence curves

```{r}

plot(
  cutTimes, cumIncTreated_de, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of prostate cancer death"),
    paste("under elimination of competing events")
  )
)
lines(
  cutTimes,
  cumIncPlacebo_de,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2),
  lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)
```

## Total effect estimation.

For this effect, we will only use the unstabilized weights we calculated for censoring due to loss to follow-up `weights_cens`, but using `weights_cens_stab` is possible as well.

Following equation (40) from [Young et al. Statistics in Medicine. 2020](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471).

```{r}

### Calculate hazards and cumulative incidence for the treated arm

Treated_te_hazardP <-
  nonParametricCumHaz(
    weights_cens,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = TRUE)

Treated_te_hazardO <-
  nonParametricCumHaz(
    weights_cens,
    inputdata = longProstRed,
    grp = 1,
    outcomeProstate = FALSE)

cumIncTreated_te <-
  nonParametricCumInc(Treated_te_hazardP, Treated_te_hazardO)

### Calculate cumulative hazards and cumulative incidence for the placebo arm

Placebo_te_hazardP <-
  nonParametricCumHaz(
    weights_cens,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = TRUE)

Placebo_te_hazardO <-
  nonParametricCumHaz(
    weights_cens,
    inputdata = longProstRed,
    grp = 0,
    outcomeProstate = FALSE)

cumIncPlacebo_te <-
  nonParametricCumInc(Placebo_te_hazardP, Placebo_te_hazardO)

```

- Risks under each treatment arm at 60 months of follow-up, risk difference and risk ratio

```{r}
cumIncTreated_te_60 <- cumIncTreated_te[length(cutTimes)] 

cumIncPlacebo_te_60 <- cumIncPlacebo_te[length(cutTimes)]

te_rr <- cumIncTreated_te_60 / cumIncPlacebo_te_60

te_rd <- cumIncTreated_te_60 - cumIncPlacebo_te_60

results_te <- data.frame(cumIncTreated_te_60, cumIncPlacebo_te_60, te_rr, te_rd)

knitr::kable(round(results_te, 2))
```

- Cumulative incidence curves

```{r}
plot(
  cutTimes, cumIncTreated_te, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of prostate cancer death"),
    paste("without elimination of competing events")
  )
)
lines(
  cutTimes,
  cumIncPlacebo_te,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2), lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)

```

## Total effect on death for other causes

We will use the hazards obtained for the total effect, from the previous section.

```{r}
cumIncTreated_te_comp <-
  nonParametricCumInc(Treated_te_hazardO, Treated_te_hazardP, competing = TRUE)

cumIncPlacebo_te_comp <-
  nonParametricCumInc(Placebo_te_hazardO, Placebo_te_hazardP, competing = TRUE)

```

- Risks under each treatment arm at 60 months of follow-up, risk difference and risk ratio

```{r}
cumIncTreated_te_comp_60 <- cumIncTreated_te_comp[length(cutTimes)] 

cumIncPlacebo_te_comp_60 <- cumIncPlacebo_te_comp[length(cutTimes)]

te_comp_rr <- cumIncTreated_te_comp_60 / cumIncPlacebo_te_comp_60

te_comp_rd <- cumIncTreated_te_comp_60 - cumIncPlacebo_te_comp_60

results_te_comp <- data.frame(cumIncTreated_te_comp_60, cumIncPlacebo_te_comp_60, te_comp_rr, te_comp_rd)

knitr::kable(round(results_te_comp, 2))

```

- Cumulative incidence curves

```{r}
plot(
  cutTimes,
  cumIncTreated_te_comp, type = "s",
  ylim = c(0, 1), ylab = "Risk", xlab = "Month",
  lty = 1, lwd = 1, xlim = c(0, 60),
  cex.lab = 1.2, cex.main = 1.2,
  main = c(
    paste("Risk of other causes of death (the competing event)")
  ))

lines(
  cutTimes,
  cumIncPlacebo_te_comp,
  type = "s", col = 2, ylim = c(0, 1), lwd = 1)

legend(
  "topleft", c("Treatment", "Placebo"),
  col = c(1, 2, 1, 2), lty = c(1, 1, 2, 2),
  cex = 1.2, pt.cex = 1.2, lwd = 2)

```

## Separable effects

Following equation (11) from [Stensrud et al. Journal of the American Statistical Association. 2020](https://www.tandfonline.com/doi/full/10.1080/01621459.2020.1765783)

To this matter, we will use the long format data set prepared before: `longProstRed`. 

- Make one extra copy of the exposure `rx`, named `Orx`, to be used in the hazard of death due to other causes.

```{r}
longProstRed$Orx <- longProstRed$rx
```

- Fit conditional pooled logistic regression models

```{r}
plrFitP <-
    glm(
      prostateDeath ~ rx*(dtime + I(dtime ^ 2)+ I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
      data = longProstRed,
      family = binomial()
    )
    
plrFitO <-
  glm(
    otherDeath ~ Orx * (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat + hx + hgBinary,
    data = longProstRed,
    family = binomial()
  )
```
  
- Create three copies of the data. Data set construction where Ay and Ad = 1

```{r}
treated <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time

treated$dtime <- rep(cutTimes, n)
  
treated$rx <- 1

treated$Orx <- 1
  
```

- Data set construction where Ay and Ad = 0

```{r}
placebo <-
    baseline[rep(1:n, each = length(cutTimes)), ] #One row for each time
  
placebo$dtime <- rep(cutTimes, n)

placebo$rx <- 0

placebo$Orx <- 0

```

- Dataset construction where Ay = 1, Ad = 0

```{r}
treatAy <-
    baseline[rep(1:n, each = length(cutTimes)), ]
  
treatAy$dtime <- rep(cutTimes, n)
  
treatAy$rx <- 1
  
treatAy$Orx <- 0
```

- Calculate hazards for each time interval in the three cloned data sets

```{r}
## For Ay = 1 and Ad = 1

treated$hazardP <-
    predict(plrFitP, newdata = treated, type = 'response')
  
treated$hazardO <-
    predict(plrFitO, newdata = treated, type = 'response')
  
treated$s <- (1 - treated$hazardP) * (1 - treated$hazardO)

## For Ay = 0, Ad = 0

placebo$hazardP <-
    predict(plrFitP, newdata = placebo, type = 'response')

placebo$hazardO <-
    predict(plrFitO, newdata = placebo, type = 'response')

placebo$s <- (1 - placebo$hazardP) * (1 - placebo$hazardO)

## For Ay = 1, Ad = 0

treatAy$hazardP <-
    predict(plrFitP, newdata = treatAy, type = 'response')
  
treatAy$hazardO <-
    predict(plrFitO, newdata = treatAy, type = 'response')
  
treatAy$s <- (1 - treatAy$hazardP) * (1 - treatAy$hazardO)

```

- Calculate weights for censoring for lost to follow-up

```{r}
plrFitC <-
  glm(
    eventCens ~ Orx* (dtime + I(dtime ^ 2) + I(dtime ^ 3)) + normalAct + ageCat +
      hx + hgBinary,
    data = longProstRed,
    family = binomial()
  )
  
predC <- 1 - predict(plrFitC, newdata = longProstRed, type = 'response')

# Create a new data frame with predicted hazard estimates from treated and 
weightFrame <-
  data.frame(
    dtime = treated$dtime,
    patno = treated$patno,
    predOa1 = (1 - placebo$hazardO),
    predOa0 = (1 - treated$hazardO)
  )

compWeightsO <-
  merge(longProstRed,
        weightFrame,
        by = c('patno', 'dtime'),
        all.x = TRUE, sort = FALSE)

predOa0 <- compWeightsO$predOa0

predOa1 <- compWeightsO$predOa1

cum_pred_O_0 <-
  unlist(aggregate(predOa0 ~ compWeightsO$patno, FUN = cumprod)$predOa0,
         use.names = F) #unlist because aggregate creates list

cum_pred_O_1 <-
  unlist(aggregate(predOa1 ~ compWeightsO$patno, FUN = cumprod)$predOa1,
         use.names = F) #unlist because aggregate creates list

cum_pred_C <-
  unlist(aggregate(predC ~ compWeightsO$patno, FUN = cumprod)$predC,
         use.names = F) #unlist because aggregate creates list

ipw_d <- rep(NA, length.out = length(cum_pred_C))

# index time zeros

t0 <- longProstRed$prostateDeath == 0

ipw_d <- cum_pred_O_1 / cum_pred_O_0

ipw_d <- cum_pred_O_1 / cum_pred_O_0

ipw_cens <- 1 / cum_pred_C

ipw_sep_eff <- ipw_d / cum_pred_C

cumIncTreatAyIPW <-
  discrete_cuminc_prost(ipw_sep_eff, longProstRed, grp = 1, follow_up = cutTimes)

cumIncTreatedIPW <-
  discrete_cuminc_prost(ipw_cens, longProstRed, grp = 1, follow_up = cutTimes)

cumIncPlaceboIPW <-
  discrete_cuminc_prost(ipw_cens, longProstRed, grp = 0, follow_up = cutTimes)



```

```{r}
estimatesFullData <-
  data.frame(cumIncTreatedIPW, cumIncTreatAyIPW, cumIncPlaceboIPW)

knitr::kable(round(estimatesFullData[36,], 2))
```


- Cumulative incidence curves

```{r}
plot(cutTimes,cumIncTreatedIPW, type="s",ylim=c(0,1), ylab="Cumulative incidence", 
     xlab="Months",lty=1,lwd=1,xlim=c(0,52))
lines(cutTimes,cumIncPlaceboIPW, type="s",col=2,ylim=c(0,1),lwd=1,lty=1)
lines(cutTimes,cumIncTreatAyIPW, type="s",col=3,ylim=c(0,1),lwd=1,lty=1)
legend(x=0,y=1, c(expression('Pr('*'Y'['k+1']^{"a"*'=1'} *'=1' * ')' ),expression('Pr('*'Y'['k+1']^{"a"['Y']*'=1' * ' , ' * "a"['D']*'=0'} *'=1' * ')' ),expression('Pr('*'Y'['k+1']^{"a"*'=0'} *'=1' * ')' )),bty = "n",
       col=c(1,3,2), lty=c(1,1,1),cex=1,pt.cex=1,lwd=1)
```


### [Go to: Data cleaning and setup here](https://palolili23.github.io/SER_competing_events_workshop/R/index.html)

### [Go to: Direct, total and separable effects, using G-formula](https://palolili23.github.io/SER_competing_events_workshop//R/gform.html)

### [Go to github repository](https://github.com/palolili23/SER_competing_events_workshop/)

# References

- Young JG, Stensrud MJ, Tchetgen Tchetgen EJ, Hernán MA. A causal framework for classical statistical estimands in failure time settings with competing events. Statistics in Medicine 2020. [Paper link](https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.8471) [Code](https://github.com/CausalInference/CompetingEvents_Young_SIM_2020)

- Stensrud MJ, Young JG, Didelez V, Robins JM, Hernán MA. Separable Effects for Causal Inference in the Presence of Competing Events. Journal of the American Statistical Association 2022. [Paper link](https://www.tandfonline.com/doi/full/10.1080/01621459.2020.1765783)

- Rojas-Saunero LP, Young JG, Didelez V, Ikram MA, Swanson SA. Considering questions before methods in dementia research with competing events and causal goals. American Journal of Epidemiology. 2023. [Paper link](https://academic.oup.com/aje/advance-article/doi/10.1093/aje/kwad090/7120022) [Code](https://github.com/palolili23/competing_risks_dementia)
